---
title: "Local Gyrification Index"
subtitle: "Exploratory Analyses"
author: "Zachary Heffernan"
date: today

format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true

execute: 
  echo: true
  warning: false
  message: false
  eval: true
---

```{r}
#| label: set up
#| include: false
# Import Libraries
library(tidyverse)
library(janitor)
library(mgcv)
library(kableExtra)
library(ComplexUpset)

# Source
source("R/plotting.R")
source("R/reporting.R")

# Define Paths
dir_dataset <- file.path(
  "/", "data", "secure", "bit_repo", "projects_repo", "ExtraLong", "data",
  "dataset_pnc_cross"
)

dir_evolpsy <- file.path(
  "/", "data", "secure", "bit_repo", "pi_repo", "raquel_gur", "evolpsy", "data"
)

path_imaging <- file.path(dir_dataset, "imaging_data.csv")
path_sub_ses <- file.path(dir_dataset, "subjects_sessions.csv")

path_merge <- file.path(
  dir_evolpsy, "go1_files", "pncgo1_data_merge_mar2025.csv"
)

path_env_ses <- file.path(dir_evolpsy, "zac_data", "env_scores.csv")
path_env_trauma <- file.path(dir_evolpsy, "zac_data", "env_for_zac.csv")

path_goassess <- file.path(
  "/", "data", "secure", "bit_repo", "projects_repo", "goassess_scoring",
  "results", "data_score.csv"
)

# Read Data
raw_imaging <- read_csv(path_imaging)
raw_sub_ses <- read_csv(path_sub_ses)

raw_merge <- read_csv(path_merge)

raw_env_ses <- read_csv(path_env_ses)
raw_env_trauma <- read_csv(path_env_trauma)

raw_goassess <- read_csv(path_goassess)
```

```{r}
#| label: clean data
#| include: false
# Demographics
demo <- raw_sub_ses |>
  select(bblid, age) |>
  left_join(
    select(raw_merge, bblid, sex, clinical_goassessDxpmr7),
    by = "bblid"
  ) |>
  rename(dx = clinical_goassessDxpmr7) |>
  mutate(
    age = age / 12,
    dx = str_to_lower(dx),
    sex = factor(
      case_when(sex == 1 ~ "Male", sex == 2 ~ "Female"),
      levels = c("Female", "Male")
    )
  )

# Environment
env <- full_join(raw_env_ses, raw_env_trauma, by = "bblid") |>
  select(bblid, envSES, Trauma_Exposure) |>
  clean_names()

# Imaging
imaging_qc <- raw_sub_ses |>
  mutate(qc = cume_dist(euler)) |>
  filter(qc > 0.01) |>
  select(bblid, scanid)

imaging <- raw_imaging |>
  filter(atlas == "desikan_killiany", !is.na(lobe)) |>
  semi_join(imaging_qc, by = c("bblid", "scanid")) |>
  pivot_wider(values_from = value, names_from = metric) |>
  select(bblid, scanid, laterality, region, lgi, area)

global_lgi <- imaging |>
  group_by(bblid, scanid) |>
  summarize(lgi = weighted.mean(lgi, area), .groups = "drop")

# Clinical
clinical <- raw_merge |>
  select(bblid, starts_with("clinical_longfactor_corrTraits_")) |>
  pivot_longer(cols = -bblid, values_to = "score", names_to = "factor") |>
  mutate(factor = str_replace(factor, "clinical_longfactor_corrTraits_", ""))

# Cognitive
cnb <- raw_merge |>
  select(bblid, starts_with("cnb") & ends_with("efficiency_agereg")) |>
  pivot_longer(cols = -bblid, values_to = "efficiency", names_to = "domain") |>
  mutate(
    domain = str_replace_all(domain, c("cnb_" = "", "_efficiency_agereg" = ""))
  ) |>
  filter(domain != "discounting")

# GOASSESS
goassess <-
  raw_goassess |>
  semi_join(demo, by = c("subject_id" = "bblid")) |>
  filter(version == 1) |>
  group_by(subject_id) |>
  summarize(across(starts_with("smry_"), \(x) max(x, na.rm = TRUE)))
```

# Sample Description
## Demographics Table
```{r}
#| label: demos table

demo_table <- demo |>
  summarize(
    n = n(),
    age_mean = mean(age, na.rm = TRUE),
    age_sd = sd(age, na.rm = TRUE),
    male_n = sum(sex == "Male"),
    female_n = sum(sex == "Female"),
    dx_td_n = sum(dx == "td", na.rm = TRUE),
    dx_ps_n = sum(dx == "ps", na.rm = TRUE),
    dx_op_n = sum(dx == "op", na.rm = TRUE)
  ) |>
  mutate(
    age = sprintf("%.2f (%.2f)", age_mean, age_sd),
    male = sprintf("%d (%.2f%%)", male_n, (male_n * 100 / n)),
    female = sprintf("%d (%.2f%%)", female_n, (female_n * 100 / n)),
    td = sprintf("%d (%.2f%%)", dx_td_n, (dx_td_n * 100 / n)),
    ps = sprintf("%d (%.2f%%)", dx_ps_n, (dx_ps_n * 100 / n)),
    op = sprintf("%d (%.2f%%)", dx_op_n, (dx_op_n * 100 / n))
  ) |>
  mutate(n = as.character(n)) |>
  select(n, age, male, female, td, ps, op) |>
  pivot_longer(
    cols = everything(),
    names_to = "Characteristic",
    values_to = "Value"
  ) |>
  mutate(Category = c("n", "age", "sex", "sex", "dx", "dx", "dx"))

row_hierarchy <- demo_table |>
  mutate(row_number = row_number()) |>
  group_by(Category) |>
  summarise(
    start = min(row_number),
    end = max(row_number),
    .groups = "drop"
  ) |>
  filter(start != end)

demo_kable <- demo_table |>
  select(-Category) |>
  kbl(
    caption = "PNC Cross Sectional Demographics",
    format = "html",
    col.names = c("", "Overall"),
    align = c("l", "c")
  ) |>
  kable_styling(
    full_width = FALSE,
    position = "left",
    font_size = 12
  )

for (i in seq_len(nrow(row_hierarchy))) {
  demo_kable <- demo_kable |>
    group_rows(
      group = row_hierarchy$Category[i],
      start_row = row_hierarchy$start[i],
      end_row = row_hierarchy$end[i]
    )
}

demo_kable
```

## Comorbidities
```{r}
comorbid <- goassess |>
  select(subject_id, ends_with("cat")) |>
  rename(bblid = subject_id) |>
  rename_with(\(x) str_replace_all(x, c("smry_" = "", "_cat" = ""))) |>
  mutate(across(
    -bblid,
    \(x) case_when(x %in% c(0, 1, 2, 3) ~ FALSE, x == 4 ~ TRUE)
  )) |>
  left_join(global_lgi, by = "bblid") |>
  select(-bblid, -scanid) |>
  drop_na()

sets <- setdiff(names(comorbid), "lgi")

psy_intersections <- unlist(
  lapply(
    seq_along(sets),
    \(n) combn(sets, n, simplify = FALSE)
  ),
  recursive = FALSE
) |>
  keep(\(x) "psy" %in% x)

psy_queries <- map(
  psy_intersections,
  \(x) {
    upset_query(
      intersect = x,
      color = "red",
      fill = "red"
    )
  }
)

plot_comorbid <- upset(
  comorbid,
  sets,
  base_annotations = list(
    "Intersection size" = intersection_size(
      counts = FALSE
    )
  ),
  annotations = list(
    "LGI" = ggplot(mapping = aes(x = intersection, y = lgi, color = psy)) +
      geom_boxplot() +
      scale_color_manual(
        values = c(`TRUE` = "red", `FALSE` = "black"),
        guide = "none"
      )
  ),
  queries = psy_queries
)

plot_comorbid
```

# Main Effects
  - [There is a significant nonlinear effect of Age on Global LGI, where Global LGI declines with Age.](#main-age)
  - [There is a significant positive correlation between Socioeconomic Status and Global LGI.](#main-ses)
  - [There is a significant nonlinear effect of Trauma Exposure on Global LGI, where those with higher Trauma Exposure have lower Global LGI.](#main-trauma)
  - [There is a significant difference in Global LGI between sexes, with Males having higher Global LGI than Females.](#main-sex)
  - [There is a significant decreasing trend in Global LGI across increasing levels of GOASSESS Psychosis Scores.](#main-goa-psy)

## Global LGI ~ Age {#main-age}
Global LGI is significantly negatively correlated with age. This effect is significantly nonlinear, though the nonlinearity is subtle.
```{r}
#| label: lgi ~ age
data_lgi_age <-
  inner_join(global_lgi, demo, by = c("bblid")) |>
  select(lgi, age) |>
  drop_na()

test_lgi_age <-
  cor.test(data_lgi_age$lgi, data_lgi_age$age)

plot_lgi_age <-
  ggplot(data = data_lgi_age, aes(x = age, y = lgi)) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Correlation of Global LGI and Age",
    subtitle = report_correlation(test_lgi_age),
    x = "Age",
    y = "LGI (global)"
  )

plot_lgi_age
```

### Global LGI ~ s(Age)
```{r}
#| label: lgi ~ s(age)
model_lgi_age <- gam(
  lgi ~ age + s(age, m = c(2, 0)),
  data = data_lgi_age,
  method = "REML"
)

summary(model_lgi_age)

plot_lgi_age_linearity <- ggplot(data = data_lgi_age, aes(x = age, y = lgi)) +
  geom_point() +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = FALSE,
    aes(color = "Linear")
  ) +
  geom_smooth(
    method = "gam",
    formula = y ~ s(x),
    se = FALSE,
    aes(color = "Nonlinear")
  ) +
  scale_color_manual(
    values = c("Linear" = "blue", "Nonlinear" = "red"),
    name = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(
    title = "Linearity of Relationship Between Global LGI and Age",
    x = "Age",
    y = "LGI (global)"
  )

plot_lgi_age_linearity
```

## Global LGI ~ Socioeconomic Status {#main-ses}
Global LGI is signitifantly positively correlated with socioeconomic status. This effect is not significantly nonlinear.
```{r}
#| label: lgi ~ ses
data_lgi_ses <-
  inner_join(global_lgi, env, by = "bblid") |>
  select(lgi, env_ses) |>
  drop_na()

test_lgi_ses <-
  cor.test(data_lgi_ses$lgi, data_lgi_ses$env_ses)

plot_lgi_ses <-
  ggplot(data = data_lgi_ses, aes(x = env_ses, y = lgi)) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Correlation of Global LGI and Socioeconomic Status",
    subtitle = report_correlation(test_lgi_ses),
    x = "Socioeconomic Status",
    y = "LGI (global)"
  )

plot_lgi_ses
```

### Global LGI ~ s(Socioeconomic Status)
```{r}
#| label: lgi ~ s(ses)
model_lgi_ses <- gam(
  lgi ~ env_ses + s(env_ses, m = c(2, 0)),
  data = data_lgi_ses,
  method = "REML"
)

summary(model_lgi_ses)
```

## Global LGI ~ Trauma Exposure {#main-trauma}
Global LGI is significantly negatively correlated with Trauma Exposure. This effect is significantly nonlinear.
```{r}
#| label: lgi ~ trauma
data_lgi_trauma <-
  inner_join(global_lgi, env, by = "bblid") |>
  select(lgi, trauma_exposure) |>
  drop_na()

test_lgi_trauma <-
  cor.test(data_lgi_trauma$lgi, data_lgi_trauma$trauma_exposure)

plot_lgi_trauma <-
  ggplot(data = data_lgi_trauma, aes(x = trauma_exposure, y = lgi)) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Correlation of Global LGI and Trauma Exposure",
    subtitle = report_correlation(test_lgi_trauma),
    x = "Trauma Exposure",
    y = "LGI (global)"
  )

plot_lgi_trauma
```

### Global LGI ~ s(Trauma Exposure)
```{r}
#| label: lgi ~ s(trauma)
model_lgi_trauma <- gam(
  lgi ~ trauma_exposure + s(trauma_exposure, k = 7, m = c(2, 0)),
  data = data_lgi_trauma,
  method = "REML"
)

summary(model_lgi_trauma)

plot_lgi_trauma_linearity <-
  ggplot(data = data_lgi_trauma, aes(x = trauma_exposure, y = lgi)) +
  geom_point() +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = FALSE,
    aes(color = "Linear")
  ) +
  geom_smooth(
    method = "gam",
    formula = y ~ s(x, k = 7),
    se = FALSE,
    aes(color = "Nonlinear")
  ) +
  scale_color_manual(
    values = c("Linear" = "blue", "Nonlinear" = "red"),
    name = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(
    title = "Linearity of Relationship Between Global LGI and Trauma Exposure",
    x = "Trauma Exposure",
    y = "LGI (global)"
  )

plot_lgi_trauma_linearity
```

## Global LGI ~ Sex {#main-sex}
There are significant differences in Global LGI between sexes, with Males having higher Global LGI than Females.
```{r}
#| label: lgi ~ sex
data_lgi_sex <-
  inner_join(global_lgi, demo, by = "bblid") |>
  select(lgi, sex) |>
  drop_na()

test_lgi_sex <-
  t.test(lgi ~ sex, data = data_lgi_sex)

plot_lgi_sex <-
  ggplot(data = data_lgi_sex, aes(x = sex, y = lgi, color = sex, fill = sex)) +
  geom_jitter(alpha = 0.25) +
  geom_violin(alpha = 0.5) +
  theme_minimal() +
  theme(legend.position = "none") +
  color_scale_sex +
  fill_scale_sex +
  labs(
    title = "Differences in Global LGI between Sexes",
    subtitle = report_t_test(test_lgi_sex),
    x = "Sex",
    y = "LGI (global)"
  )

plot_lgi_sex
```

## Global LGI ~ GOASSESS Psychosis Category {#main-goa-psy}
Global LGI shows a significant decreasing trend across increasing levels of GOASSESS Psychosis Scores.
```{r}
#| label: lgi ~ goa_smry_psy
data_lgi_psy <-
  inner_join(global_lgi, goassess, by = c("bblid" = "subject_id")) |>
  select(lgi, smry_psy_cat) |>
  filter(smry_psy_cat %in% c(0, 1, 2, 3, 4)) |>
  mutate(smry_psy_cat = ordered(smry_psy_cat)) |>
  drop_na()

test_lgi_psy <-
  clinfun::jonckheere.test(
    x = data_lgi_psy$lgi,
    g = data_lgi_psy$smry_psy_cat,
    alternative = "decreasing"
  )

plot_lgi_psy <-
  ggplot(
    data = data_lgi_psy,
    aes(x = smry_psy_cat, y = lgi, color = smry_psy_cat, fill = smry_psy_cat)
  ) +
  geom_jitter(alpha = 0.25) +
  geom_violin(alpha = 0.5) +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_x_discrete(
    labels = c(
      "0" = "Not Endorsed",
      "1" = "Endorsed",
      "2" = "Subthreshold",
      "3" = "Threshold",
      "4" = "Significant"
    )
  ) +
  labs(
    title = "Differences in Global LGI between Psychosis Categories",
    subtitle = report_jt(test_lgi_psy),
    x = "GOASSESS Psychosis Category",
    y = "LGI (global)"
  )

plot_lgi_psy
```

# Two-Way Interactions
  - [There are no significant linear or nonlinear interactions between age and sex. The most complex model supported is `lgi ~ s(age) + sex`.](#two-age-sex)

## Global LGI ~ Age * Sex {#two-age-sex}
```{r}
#| label: lgi ~ age * sex
data_lgi_age_sex <-
  inner_join(global_lgi, demo, by = c("bblid")) |>
  mutate(
    sex_ord = ordered(sex, levels = c("Female", "Male")),
    age_c = age - mean(age)
  ) |>
  select(lgi, age, age_c, sex, sex_ord) |>
  drop_na()
```

### Global LGI ~ Age * Sex (Final)
```{r}
#| label: lgi ~ age * sex [final]
model_lgi_age_sex_final <- gam(
  lgi ~ sex + s(age),
  data = data_lgi_age_sex,
  method = "REML"
)

prediction_lgi_age_sex <- data_lgi_age_sex |>
  group_by(sex) |>
  summarise(
    age = list(
      seq(
        min(age, na.rm = TRUE),
        max(age, na.rm = TRUE),
        length.out = 250
      )
    ),
    .groups = "drop"
  ) |>
  unnest(age)

fit_data_lgi_age_sex <- predict(
  model_lgi_age_sex_final,
  newdata = prediction_lgi_age_sex,
  type = "response",
  se.fit = TRUE
)

prediction_lgi_age_sex <- prediction_lgi_age_sex |>
  mutate(
    fit = as.vector(pluck(fit_data_lgi_age_sex, "fit")),
    se_fit = as.vector(pluck(fit_data_lgi_age_sex, "se.fit")),
    lower = fit - 1.96 * se_fit,
    upper = fit + 1.96 * se_fit
  ) |>
  select(age, sex, fit, upper, lower)

plot_lgi_age_sex <-
  ggplot(mapping = aes(x = age, y = lgi, color = sex, fill = sex)) +
  geom_point(data = data_lgi_age_sex, alpha = 0.25, size = 1) +
  geom_line(
    data = prediction_lgi_age_sex,
    mapping = aes(y = fit),
    linewidth = 0.75
  ) +
  geom_ribbon(
    data = prediction_lgi_age_sex,
    mapping = aes(y = fit, ymin = lower, ymax = upper),
    alpha = 0.5,
    color = NA
  ) +
  theme_linedraw() +
  theme(legend.position = "bottom") +
  color_scale_sex +
  fill_scale_sex +
  labs(
    title = "LGI Developmental Trajectories by Sex",
    subtitle = "lgi ~ s(age) + sex",
    x = "Age",
    y = "LGI (global)"
  )

plot_lgi_age_sex
```

### Global LGI ~ Age + Sex + s(Age, m = c(2,0))
Sex contributes when accounting for age. Age contributes when accounting for sex, and this effect is significantly nonlinear.
```{r}
#| label: lgi ~ age + sex + s(age, m = 0)
model_lgi_age_sex_1 <- gam(
  lgi ~ age_c + sex + s(age_c, m = c(2, 0)),
  data = data_lgi_age_sex,
  method = "REML"
)

summary(model_lgi_age_sex_1)
```

### Global LGI ~ Sex * Age + s(Age, m = c(2,0))
Age and sex do not have a significant linear interaction.
```{r}
#| label: lgi ~ age * sex + s(age, m = 0)
model_lgi_age_sex_2 <- gam(
  lgi ~ age_c * sex + s(age_c, m = c(2, 0)),
  data = data_lgi_age_sex,
  method = "REML"
)

summary(model_lgi_age_sex_2)
```

### Global LGI ~ Sex * Age + s(Age, m = c(2,0)) + s(Age, m = c(2,0), by = Sex)
Age and sex do not have a significant nonlinear interaction.
```{r}
#| label: lgi ~ age * sex + s(age, m = 0) + s(age, m = 0, by = sex)
model_lgi_age_sex_3 <- gam(
  lgi ~ age_c * sex +
    s(age_c, m = c(2, 0)) +
    s(age_c, m = c(2, 0), by = sex_ord),
  data = data_lgi_age_sex,
  method = "REML"
)

summary(model_lgi_age_sex_3)
```
