---
title: "Local Gyrification Index"
subtitle: "Exploratory Analyses"
author: "Zachary Heffernan"
date: today

format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true

execute: 
  echo: true
  warning: false
  message: false
  eval: true
---

```{r}
#| label: set up
#| include: false
# Import Libraries
library(tidyverse)
library(janitor)
library(mgcv)
library(kableExtra)
library(ComplexUpset)

# Source
source("R/plotting.R")
source("R/reporting.R")

# Define Paths
dir_dataset <- file.path(
  "/", "data", "secure", "bit_repo", "projects_repo", "ExtraLong", "data",
  "dataset_pnc_cross"
)

dir_evolpsy <- file.path(
  "/", "data", "secure", "bit_repo", "pi_repo", "raquel_gur", "evolpsy", "data"
)

path_imaging <- file.path(dir_dataset, "imaging_data.csv")
path_sub_ses <- file.path(dir_dataset, "subjects_sessions.csv")

path_merge <- file.path(
  dir_evolpsy, "go1_files", "pncgo1_data_merge_mar2025.csv"
)

path_env_ses <- file.path(dir_evolpsy, "zac_data", "env_scores.csv")
path_env_trauma <- file.path(dir_evolpsy, "zac_data", "env_for_zac.csv")

path_goassess <- file.path(
  "/", "data", "secure", "bit_repo", "projects_repo", "goassess_scoring",
  "results", "data_score.csv"
)

# Read Data
raw_imaging <- read_csv(path_imaging)
raw_sub_ses <- read_csv(path_sub_ses)

raw_merge <- read_csv(path_merge)

raw_env_ses <- read_csv(path_env_ses)
raw_env_trauma <- read_csv(path_env_trauma)

raw_goassess <- read_csv(path_goassess)
```

```{r}
#| label: clean data
#| include: false
# Demographics
demo <- raw_sub_ses |>
  select(bblid, age) |>
  left_join(
    select(raw_merge, bblid, sex, clinical_goassessDxpmr7),
    by = "bblid"
  ) |>
  rename(dx = clinical_goassessDxpmr7) |>
  mutate(
    age = age / 12,
    dx = str_to_lower(dx),
    sex = factor(
      case_when(sex == 1 ~ "Male", sex == 2 ~ "Female"),
      levels = c("Female", "Male")
    )
  )

# Environment
env <- full_join(raw_env_ses, raw_env_trauma, by = "bblid") |>
  select(bblid, envSES, Trauma_Exposure) |>
  clean_names()

# Imaging
imaging_qc <- raw_sub_ses |>
  mutate(qc = cume_dist(euler)) |>
  filter(qc > 0.01) |>
  select(bblid, scanid)

imaging <- raw_imaging |>
  filter(atlas == "desikan_killiany", !is.na(lobe)) |>
  semi_join(imaging_qc, by = c("bblid", "scanid")) |>
  pivot_wider(values_from = value, names_from = metric) |>
  select(bblid, scanid, laterality, region, lgi, area)

global_lgi <- imaging |>
  group_by(bblid, scanid) |>
  summarize(lgi = weighted.mean(lgi, area), .groups = "drop")

# Clinical
clinical <- raw_merge |>
  select(bblid, starts_with("clinical_longfactor_corrTraits_")) |>
  pivot_longer(cols = -bblid, values_to = "score", names_to = "factor") |>
  mutate(factor = str_replace(factor, "clinical_longfactor_corrTraits_", ""))

# Cognitive
cnb <- raw_merge |>
  select(bblid, starts_with("cnb") & ends_with("efficiency_agereg")) |>
  pivot_longer(cols = -bblid, values_to = "efficiency", names_to = "domain") |>
  mutate(
    domain = str_replace_all(domain, c("cnb_" = "", "_efficiency_agereg" = ""))
  ) |>
  filter(domain != "discounting")

# GOASSESS
goassess <-
  raw_goassess |>
  semi_join(demo, by = c("subject_id" = "bblid")) |>
  filter(version == 1) |>
  group_by(subject_id) |>
  summarize(across(starts_with("smry_"), \(x) max(x, na.rm = TRUE)))
```

# Sample Description
## Demographics Table
```{r}
#| label: demos table

demo_table <- demo |>
  summarize(
    n = n(),
    age_mean = mean(age, na.rm = TRUE),
    age_sd = sd(age, na.rm = TRUE),
    male_n = sum(sex == "Male"),
    female_n = sum(sex == "Female"),
    dx_td_n = sum(dx == "td", na.rm = TRUE),
    dx_ps_n = sum(dx == "ps", na.rm = TRUE),
    dx_op_n = sum(dx == "op", na.rm = TRUE)
  ) |>
  mutate(
    age = sprintf("%.2f (%.2f)", age_mean, age_sd),
    male = sprintf("%d (%.2f%%)", male_n, (male_n * 100 / n)),
    female = sprintf("%d (%.2f%%)", female_n, (female_n * 100 / n)),
    td = sprintf("%d (%.2f%%)", dx_td_n, (dx_td_n * 100 / n)),
    ps = sprintf("%d (%.2f%%)", dx_ps_n, (dx_ps_n * 100 / n)),
    op = sprintf("%d (%.2f%%)", dx_op_n, (dx_op_n * 100 / n))
  ) |>
  mutate(n = as.character(n)) |>
  select(n, age, male, female, td, ps, op) |>
  pivot_longer(
    cols = everything(),
    names_to = "Characteristic",
    values_to = "Value"
  ) |>
  mutate(Category = c("n", "age", "sex", "sex", "dx", "dx", "dx"))

row_hierarchy <- demo_table |>
  mutate(row_number = row_number()) |>
  group_by(Category) |>
  summarise(
    start = min(row_number),
    end = max(row_number),
    .groups = "drop"
  ) |>
  filter(start != end)

demo_kable <- demo_table |>
  select(-Category) |>
  kbl(
    caption = "PNC Cross Sectional Demographics",
    format = "html",
    col.names = c("", "Overall"),
    align = c("l", "c")
  ) |>
  kable_styling(
    full_width = FALSE,
    position = "left",
    font_size = 12
  )

for (i in seq_len(nrow(row_hierarchy))) {
  demo_kable <- demo_kable |>
    group_rows(
      group = row_hierarchy$Category[i],
      start_row = row_hierarchy$start[i],
      end_row = row_hierarchy$end[i]
    )
}

demo_kable
```

## Comorbidities
```{r}
#| label: comorbidities
comorbid <- goassess |>
  select(subject_id, ends_with("cat")) |>
  rename(bblid = subject_id) |>
  rename_with(\(x) str_replace_all(x, c("smry_" = "", "_cat" = ""))) |>
  mutate(across(
    -bblid,
    \(x) case_when(x %in% c(0, 1, 2, 3) ~ FALSE, x == 4 ~ TRUE)
  )) |>
  left_join(global_lgi, by = "bblid") |>
  select(-bblid, -scanid) |>
  drop_na()

sets <- setdiff(names(comorbid), "lgi")

psy_intersections <- unlist(
  lapply(
    seq_along(sets),
    \(n) combn(sets, n, simplify = FALSE)
  ),
  recursive = FALSE
) |>
  keep(\(x) "psy" %in% x)

psy_queries <- map(
  psy_intersections,
  \(x) {
    upset_query(
      intersect = x,
      color = "red",
      fill = "red"
    )
  }
)

plot_comorbid <- upset(
  comorbid,
  sets,
  base_annotations = list(
    "Intersection size" = intersection_size(
      counts = FALSE
    )
  ),
  annotations = list(
    "LGI" = ggplot(mapping = aes(x = intersection, y = lgi, color = psy)) +
      geom_boxplot() +
      scale_color_manual(
        values = c(`TRUE` = "red", `FALSE` = "black"),
        guide = "none"
      )
  ),
  queries = psy_queries
)

plot_comorbid
```

# Local Gyrification Index
## LGI vs eTIV
```{r}
#| label: lgi vs etiv
data_etiv <- raw_imaging |>
  filter(region == "EstimatedTotalIntraCranialVol") |>
  rename(etiv = value) |>
  select(bblid, scanid, etiv)

data_lgi_etiv <- left_join(global_lgi, data_etiv, by = c("bblid", "scanid"))

plot_lgi_etiv <-
  ggplot(data = data_lgi_etiv, mapping = aes(x = etiv, y = lgi)) +
  geom_point() +
  theme_linedraw()

plot_lgi_etiv
```

# Main Effects
  - [There is a significant nonlinear effect of Age on Global LGI, where Global LGI declines with Age.](#main-age)
  - [There is a significant positive correlation between Socioeconomic Status and Global LGI.](#main-ses)
  - [There is a significant nonlinear effect of Trauma Exposure on Global LGI, where those with higher Trauma Exposure have lower Global LGI.](#main-trauma)
  - [There is a significant difference in Global LGI between sexes, with Males having higher Global LGI than Females.](#main-sex)
  - [There is a significant decreasing trend in Global LGI across increasing levels of GOASSESS Psychosis Scores.](#main-goa-psy)

## Global LGI ~ Age {#main-age}
Global LGI is significantly negatively correlated with age. This effect is significantly nonlinear, though the nonlinearity is subtle.
```{r}
#| label: lgi ~ age
data_lgi_age <-
  inner_join(global_lgi, demo, by = c("bblid")) |>
  select(lgi, age) |>
  drop_na()

test_lgi_age <-
  cor.test(data_lgi_age$lgi, data_lgi_age$age)

plot_lgi_age <-
  ggplot(data = data_lgi_age, aes(x = age, y = lgi)) +
  geom_point() +
  theme_linedraw() +
  labs(
    title = "Correlation of Global LGI and Age",
    subtitle = report_correlation(test_lgi_age),
    x = "Age",
    y = "LGI (global)"
  )

plot_lgi_age
```

### Global LGI ~ s(Age)
```{r}
#| label: lgi ~ s(age)
model_lgi_age <- gam(
  lgi ~ age + s(age, m = c(2, 0)),
  data = data_lgi_age,
  method = "REML"
)

summary(model_lgi_age)

plot_lgi_age_linearity <- ggplot(data = data_lgi_age, aes(x = age, y = lgi)) +
  geom_point() +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = FALSE,
    aes(color = "Linear")
  ) +
  geom_smooth(
    method = "gam",
    formula = y ~ s(x),
    se = FALSE,
    aes(color = "Nonlinear")
  ) +
  scale_color_manual(
    values = c("Linear" = "blue", "Nonlinear" = "red"),
    name = NULL
  ) +
  theme_linedraw() +
  theme(legend.position = "bottom") +
  labs(
    title = "Linearity of Relationship Between Global LGI and Age",
    x = "Age",
    y = "LGI (global)"
  )

plot_lgi_age_linearity
```

## Global LGI ~ Socioeconomic Status {#main-ses}
Global LGI is signitifantly positively correlated with socioeconomic status. This effect is not significantly nonlinear.
```{r}
#| label: lgi ~ ses
data_lgi_ses <-
  inner_join(global_lgi, env, by = "bblid") |>
  select(lgi, env_ses) |>
  drop_na()

test_lgi_ses <-
  cor.test(data_lgi_ses$lgi, data_lgi_ses$env_ses)

plot_lgi_ses <-
  ggplot(data = data_lgi_ses, aes(x = env_ses, y = lgi)) +
  geom_point() +
  theme_linedraw() +
  labs(
    title = "Correlation of Global LGI and Socioeconomic Status",
    subtitle = report_correlation(test_lgi_ses),
    x = "Socioeconomic Status",
    y = "LGI (global)"
  )

plot_lgi_ses
```

### Global LGI ~ s(Socioeconomic Status)
```{r}
#| label: lgi ~ s(ses)
model_lgi_ses <- gam(
  lgi ~ env_ses + s(env_ses, m = c(2, 0)),
  data = data_lgi_ses,
  method = "REML"
)

summary(model_lgi_ses)
```

## Global LGI ~ Trauma Exposure {#main-trauma}
Global LGI is significantly negatively correlated with Trauma Exposure. This effect is significantly nonlinear.
```{r}
#| label: lgi ~ trauma
data_lgi_trauma <-
  inner_join(global_lgi, env, by = "bblid") |>
  select(lgi, trauma_exposure) |>
  drop_na()

test_lgi_trauma <-
  cor.test(data_lgi_trauma$lgi, data_lgi_trauma$trauma_exposure)

plot_lgi_trauma <-
  ggplot(data = data_lgi_trauma, aes(x = trauma_exposure, y = lgi)) +
  geom_point() +
  theme_linedraw() +
  labs(
    title = "Correlation of Global LGI and Trauma Exposure",
    subtitle = report_correlation(test_lgi_trauma),
    x = "Trauma Exposure",
    y = "LGI (global)"
  )

plot_lgi_trauma
```

### Global LGI ~ s(Trauma Exposure)
```{r}
#| label: lgi ~ s(trauma)
model_lgi_trauma <- gam(
  lgi ~ trauma_exposure + s(trauma_exposure, k = 7, m = c(2, 0)),
  data = data_lgi_trauma,
  method = "REML"
)

summary(model_lgi_trauma)

plot_lgi_trauma_linearity <-
  ggplot(data = data_lgi_trauma, aes(x = trauma_exposure, y = lgi)) +
  geom_point() +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = FALSE,
    aes(color = "Linear")
  ) +
  geom_smooth(
    method = "gam",
    formula = y ~ s(x, k = 7),
    se = FALSE,
    aes(color = "Nonlinear")
  ) +
  scale_color_manual(
    values = c("Linear" = "blue", "Nonlinear" = "red"),
    name = NULL
  ) +
  theme_linedraw() +
  theme(legend.position = "bottom") +
  labs(
    title = "Linearity of Relationship Between Global LGI and Trauma Exposure",
    x = "Trauma Exposure",
    y = "LGI (global)"
  )

plot_lgi_trauma_linearity
```

## Global LGI ~ Sex {#main-sex}
There are significant differences in Global LGI between sexes, with Males having higher Global LGI than Females.
```{r}
#| label: lgi ~ sex
data_lgi_sex <-
  inner_join(global_lgi, demo, by = "bblid") |>
  select(lgi, sex) |>
  drop_na()

test_lgi_sex <-
  t.test(lgi ~ sex, data = data_lgi_sex)

plot_lgi_sex <-
  ggplot(data = data_lgi_sex, aes(x = sex, y = lgi, color = sex, fill = sex)) +
  geom_jitter(alpha = 0.25) +
  geom_violin(alpha = 0.5) +
  theme_linedraw() +
  theme(legend.position = "none") +
  color_scale_sex +
  fill_scale_sex +
  labs(
    title = "Differences in Global LGI between Sexes",
    subtitle = report_t_test(test_lgi_sex),
    x = "Sex",
    y = "LGI (global)"
  )

plot_lgi_sex
```

## Global LGI ~ GOASSESS Psychosis Category {#main-goa-psy}
Global LGI shows a significant decreasing trend across increasing levels of GOASSESS Psychosis Scores.
```{r}
#| label: lgi ~ goa_smry_psy
data_lgi_psy <-
  inner_join(global_lgi, goassess, by = c("bblid" = "subject_id")) |>
  select(lgi, smry_psy_cat) |>
  filter(smry_psy_cat %in% c(0, 1, 2, 3, 4)) |>
  mutate(smry_psy_cat = ordered(smry_psy_cat)) |>
  drop_na()

test_lgi_psy <-
  clinfun::jonckheere.test(
    x = data_lgi_psy$lgi,
    g = data_lgi_psy$smry_psy_cat,
    alternative = "decreasing"
  )

plot_lgi_psy <-
  ggplot(
    data = data_lgi_psy,
    aes(x = smry_psy_cat, y = lgi, color = smry_psy_cat, fill = smry_psy_cat)
  ) +
  geom_jitter(alpha = 0.25) +
  geom_violin(alpha = 0.5) +
  theme_linedraw() +
  theme(legend.position = "none") +
  scale_x_discrete(
    labels = c(
      "0" = "Not Endorsed",
      "1" = "Endorsed",
      "2" = "Subthreshold",
      "3" = "Threshold",
      "4" = "Significant"
    )
  ) +
  labs(
    title = "Differences in Global LGI between Psychosis Categories",
    subtitle = report_jt(test_lgi_psy),
    x = "GOASSESS Psychosis Category",
    y = "LGI (global)"
  )

plot_lgi_psy
```

# Two-Way Interactions
  - [Global LGI was nonlinearly associated with both age and SES, with no evidence of either a linear or nonlinear age × SES interaction. `lgi ~ s(age) + s(ses)`](#two-age-ses)
  - [Global LGI showed a significant nonlinear association with age and differed significantly by sex, but there was no evidence that the age–LGI relationship differed between females and males. `lgi ~ s(age) + sex`.](#two-age-sex)
  - [Global LGI was nonlinearly associated with age and linearly associated with trauma exposure, with a significant linear age × trauma interaction but no evidence of a more complex nonlinear interaction. `lgi ~ s(age) + truma + age:trauma`](#two-age-trauma)
  - [Global LGI differed by sex and showed a significant nonlinear association with SES, with evidence that the shape of this SES–LGI relationship differed between females and males. `lgi ~ s(ses, by = sex) + sex`.](#two-ses-sex)
  - [Greater trauma exposure was significantly associated with lower global LGI, and global LGI differed by sex, but there was no evidence that the trauma–LGI relationship differed between females and males. `lgi ~ sex + trauma`.](#two-sex-trauma)

## Global LGI ~ Age * SES {#two-age-ses}
Age and SES were not significantly correlated. When modeled together, both age and SES showed significant nonlinear associations with global LGI, indicating that each contributed independently to LGI in a nonlinear manner. There was no evidence of a significant linear age × SES interaction, and a more flexible tensor-product interaction was also not significant. The most complex model supported by the data was therefore `lgi ~ s(age) + s(ses)`.
```{r}
#| label: lgi age ses
data_lgi_age_ses <-
  global_lgi |>
  inner_join(demo, by = "bblid") |>
  inner_join(env, by = "bblid") |>
  rename(ses = env_ses) |>
  mutate(age_c = age - mean(age), ses_c = ses - mean(ses)) |>
  select(lgi, age, age_c, ses, ses_c) |>
  drop_na()

test_age_ses <-
  cor.test(data_lgi_age_ses$age, data_lgi_age_ses$ses)

test_age_ses
```

```{r}
#| label: lgi age ses 1
model_lgi_age_ses_1 <- gam(
  lgi ~ age_c +
    ses_c +
    s(age_c, m = c(2, 0)) +
    s(ses_c, k = 7, m = c(2, 0)),
  data = data_lgi_age_ses,
  method = "REML"
)

summary(model_lgi_age_ses_1)
```

```{r}
#| label: lgi age ses 2
model_lgi_age_ses_2 <- gam(
  lgi ~ age_c * ses_c +
    s(age_c, m = c(2, 0)) +
    s(ses_c, k = 7, m = c(2, 0)),
  data = data_lgi_age_ses,
  method = "REML"
)

summary(model_lgi_age_ses_2)
```

```{r}
#| label: lgi age ses 3
model_lgi_age_ses_3 <- gam(
  lgi ~ s(age) +
    s(ses, k = 7) +
    ti(age, ses, k = c(10, 7)),
  data = data_lgi_age_ses,
  method = "REML"
)

summary(model_lgi_age_ses_3)
```

```{r}
#| label: lgi age ses final
model_lgi_age_ses_final <- gam(
  lgi ~ s(age) + s(ses),
  data = data_lgi_age_ses,
  method = "REML"
)

ses_0 <- min(data_lgi_age_ses$ses, na.rm = TRUE)
ses_16 <- unname(quantile(data_lgi_age_ses$ses, probs = 1 / 6))
ses_33 <- unname(quantile(data_lgi_age_ses$ses, probs = 1 / 3))
ses_50 <- unname(quantile(data_lgi_age_ses$ses, probs = 1 / 2))
ses_66 <- unname(quantile(data_lgi_age_ses$ses, probs = 2 / 3))
ses_83 <- unname(quantile(data_lgi_age_ses$ses, probs = 5 / 6))
ses_100 <- max(data_lgi_age_ses$ses, na.rm = TRUE)

prediction_lgi_age_ses <- data_lgi_age_ses |>
  mutate(
    ses = case_when(
      ses <= ses_33 ~ ses_16,
      ses > ses_33 & ses <= ses_66 ~ ses_50,
      ses > ses_66 ~ ses_83
    )
  ) |>
  group_by(ses) |>
  summarise(
    age = list(
      seq(
        min(age, na.rm = TRUE),
        max(age, na.rm = TRUE),
        length.out = 250
      )
    ),
    .groups = "drop"
  ) |>
  unnest(age)

fit_lgi_age_ses <- predict(
  model_lgi_age_ses_final,
  newdata = prediction_lgi_age_ses,
  type = "response",
  se.fit = TRUE
)

prediction_lgi_age_ses <- prediction_lgi_age_ses |>
  mutate(
    fit = as.vector(pluck(fit_lgi_age_ses, "fit")),
    se_fit = as.vector(pluck(fit_lgi_age_ses, "se.fit")),
    lower = fit - 1.96 * se_fit,
    upper = fit + 1.96 * se_fit
  ) |>
  select(age, ses, fit, upper, lower)

plot_lgi_age_ses <-
  ggplot(
    mapping = aes(
      x = age, y = lgi, group = ses, color = ses, fill = ses
    )
  ) +
  geom_point(data = data_lgi_age_ses, alpha = 0.25, size = 1) +
  geom_line(
    data = prediction_lgi_age_ses,
    mapping = aes(y = fit),
    linewidth = 0.75
  ) +
  geom_ribbon(
    data = prediction_lgi_age_ses,
    mapping = aes(y = fit, ymin = lower, ymax = upper),
    alpha = 0.5,
    color = NA
  ) +
  scale_color_viridis_c(
    limits = c(ses_0, ses_100),
    breaks = c(ses_16, ses_50, ses_83),
    labels = scales::label_number(accuracy = 0.1)

  ) +
  scale_fill_viridis_c(
    limits = c(ses_0, ses_100),
    breaks = c(ses_16, ses_50, ses_83),
    labels = scales::label_number(accuracy = 0.1)
  ) +
  theme_linedraw() +
  theme(legend.position = "bottom") +
  labs(
    title = "LGI Developmental Trajectories by SES",
    subtitle = "lgi ~ s(age) + s(ses)",
    x = "Age",
    y = "LGI (global)",
    color = "SES",
    fill = "SES"
  )

plot_lgi_age_ses
```

## Global LGI ~ Age * Sex {#two-age-sex}
Age differed significantly by sex, with females being slightly older than males on average. When age and sex were modeled together, global LGI showed a significant nonlinear association with age and a significant main effect of sex, with males having higher global LGI than females. There was no evidence that the age–LGI relationship differed by sex in either its linear or nonlinear component, indicating that the same nonlinear age-related pattern adequately described both sexes. The most complex model supported by the data was therefore lgi ~ s(age) + sex.
```{r}
#| label: lgi ~ age * sex
data_lgi_age_sex <-
  inner_join(global_lgi, demo, by = c("bblid")) |>
  mutate(
    sex_ord = ordered(sex, levels = c("Female", "Male")),
    age_c = age - mean(age)
  ) |>
  select(lgi, age, age_c, sex, sex_ord) |>
  drop_na()

t.test(
  age ~ sex,
  data = data_lgi_age_sex,
  var.equal = TRUE
)

plot_age_sex <-
  ggplot(
    data = data_lgi_age_sex,
    mapping = aes(x = age, group = sex, color = sex, fill = sex)
  ) +
  geom_histogram(position = "identity", binwidth = 1, alpha = 0.5) +
  theme_linedraw() +
  theme(legend.position = "bottom") +
  color_scale_sex +
  fill_scale_sex +
  labs(title = "Age Distribution by Sex", x = "Age", y = "Count")

plot_age_sex
```

### Global LGI ~ Age + Sex + s(Age, m = c(2,0))
Sex contributes when accounting for age. Age contributes when accounting for sex, and this effect is significantly nonlinear.
```{r}
#| label: lgi ~ age + sex + s(age, m = 0)
model_lgi_age_sex_1 <- gam(
  lgi ~ age_c + sex + s(age_c, m = c(2, 0)),
  data = data_lgi_age_sex,
  method = "REML"
)

summary(model_lgi_age_sex_1)
```

### Global LGI ~ Sex * Age + s(Age, m = c(2,0))
Age and sex do not have a significant linear interaction.
```{r}
#| label: lgi ~ age * sex + s(age, m = 0)
model_lgi_age_sex_2 <- gam(
  lgi ~ age_c * sex + s(age_c, m = c(2, 0)),
  data = data_lgi_age_sex,
  method = "REML"
)

summary(model_lgi_age_sex_2)
```

### Global LGI ~ Sex * Age + s(Age, m = c(2,0)) + s(Age, m = c(2,0), by = Sex)
Age and sex do not have a significant nonlinear interaction.
```{r}
#| label: lgi ~ age * sex + s(age, m = 0) + s(age, m = 0, by = sex)
model_lgi_age_sex_3 <- gam(
  lgi ~ age_c * sex +
    s(age_c, m = c(2, 0)) +
    s(age_c, m = c(2, 0), by = sex_ord),
  data = data_lgi_age_sex,
  method = "REML"
)

summary(model_lgi_age_sex_3)
```

### Global LGI ~ Age * Sex (Final)
```{r}
#| label: lgi ~ age * sex [final]
model_lgi_age_sex_final <- gam(
  lgi ~ sex + s(age),
  data = data_lgi_age_sex,
  method = "REML"
)

prediction_lgi_age_sex <- data_lgi_age_sex |>
  group_by(sex) |>
  summarise(
    age = list(
      seq(
        min(age, na.rm = TRUE),
        max(age, na.rm = TRUE),
        length.out = 250
      )
    ),
    .groups = "drop"
  ) |>
  unnest(age)

fit_data_lgi_age_sex <- predict(
  model_lgi_age_sex_final,
  newdata = prediction_lgi_age_sex,
  type = "response",
  se.fit = TRUE
)

prediction_lgi_age_sex <- prediction_lgi_age_sex |>
  mutate(
    fit = as.vector(pluck(fit_data_lgi_age_sex, "fit")),
    se_fit = as.vector(pluck(fit_data_lgi_age_sex, "se.fit")),
    lower = fit - 1.96 * se_fit,
    upper = fit + 1.96 * se_fit
  ) |>
  select(age, sex, fit, upper, lower)

plot_lgi_age_sex <-
  ggplot(mapping = aes(x = age, y = lgi, color = sex, fill = sex)) +
  geom_point(data = data_lgi_age_sex, alpha = 0.25, size = 1) +
  geom_line(
    data = prediction_lgi_age_sex,
    mapping = aes(y = fit),
    linewidth = 0.75
  ) +
  geom_ribbon(
    data = prediction_lgi_age_sex,
    mapping = aes(y = fit, ymin = lower, ymax = upper),
    alpha = 0.5,
    color = NA
  ) +
  theme_linedraw() +
  theme(legend.position = "bottom") +
  color_scale_sex +
  fill_scale_sex +
  labs(
    title = "LGI Developmental Trajectories by Sex",
    subtitle = "lgi ~ s(age) + sex",
    x = "Age",
    y = "LGI (global)"
  )

plot_lgi_age_sex
```

## Global LGI ~ Age * Trauma {#two-age-trauma}
Age and trauma exposure were positively correlated. When modeled together, global LGI showed a significant nonlinear association with age and a significant linear association with trauma exposure. There was also a significant linear age × trauma interaction, indicating that the association between trauma exposure and global LGI varied with age. However, there was no evidence that this interaction required a more complex nonlinear surface. The most complex model supported by the data was therefore `lgi ~ s(age) + trauma + age:trauma`.
```{r}
#| label: lgi age trauma
data_lgi_age_trauma <-
  global_lgi |>
  inner_join(demo, by = "bblid") |>
  inner_join(env, by = "bblid") |>
  rename(trauma = trauma_exposure) |>
  mutate(age_c = age - mean(age), trauma_c = trauma - mean(trauma)) |>
  select(lgi, age, age_c, trauma, trauma_c) |>
  drop_na()

test_age_trauma <-
  cor.test(data_lgi_age_trauma$age, data_lgi_age_trauma$trauma)

test_age_trauma

plot_age_trauma <-
  ggplot(data = data_lgi_age_trauma, aes(x = age, y = trauma)) +
  geom_point() +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = FALSE
  ) +
  theme_linedraw() +
  labs(
    title = "Correlation of Age and Trauma",
    subtitle = report_correlation(test_age_trauma),
    x = "Age",
    y = "Trauma"
  )

plot_age_trauma
```

### LGI ~ s(age) + s(trauma)
```{r}
#| label: lgi age trauma 1
model_lgi_age_trauma_1 <- gam(
  lgi ~ age_c +
    trauma_c +
    s(age_c, m = c(2, 0)) +
    s(trauma_c, k = 7, m = c(2, 0)),
  data = data_lgi_age_trauma,
  method = "REML"
)

summary(model_lgi_age_trauma_1)
```

```{r}
#| label: lgi age trauma 2
model_lgi_age_trauma_2 <- gam(
  lgi ~ age_c * trauma_c +
    s(age_c, m = c(2, 0)),
  data = data_lgi_age_trauma,
  method = "REML"
)

summary(model_lgi_age_trauma_2)
```

```{r}
#| label: lgi age trauma 3
model_lgi_age_trauma_3 <- gam(
  lgi ~ s(age) +
    s(trauma, k = 7) +
    ti(age, trauma, k = c(10, 7)),
  data = data_lgi_age_trauma,
  method = "REML"
)

summary(model_lgi_age_trauma_3)
```

```{r}
#| label: lgi age trauma final
model_lgi_age_trauma_final <- gam(
  lgi ~ s(age) + trauma + age:trauma,
  data = data_lgi_age_trauma,
  method = "REML"
)

prediction_lgi_age_trauma <- data_lgi_age_trauma |>
  filter(trauma %in% c(0, 3, 6)) |>
  group_by(trauma) |>
  summarise(
    age = list(
      seq(
        min(age, na.rm = TRUE),
        max(age, na.rm = TRUE),
        length.out = 250
      )
    ),
    .groups = "drop"
  ) |>
  unnest(age)

fit_lgi_age_trauma <- predict(
  model_lgi_age_trauma_final,
  newdata = prediction_lgi_age_trauma,
  type = "response",
  se.fit = TRUE
)

prediction_lgi_age_trauma <- prediction_lgi_age_trauma |>
  mutate(
    fit = as.vector(pluck(fit_lgi_age_trauma, "fit")),
    se_fit = as.vector(pluck(fit_lgi_age_trauma, "se.fit")),
    lower = fit - 1.96 * se_fit,
    upper = fit + 1.96 * se_fit
  ) |>
  select(age, trauma, fit, upper, lower)

plot_lgi_age_trauma <-
  ggplot(
    mapping = aes(
      x = age, y = lgi, group = trauma, color = trauma, fill = trauma
    )
  ) +
  geom_point(data = data_lgi_age_trauma, alpha = 0.25, size = 1) +
  geom_line(
    data = prediction_lgi_age_trauma,
    mapping = aes(y = fit),
    linewidth = 0.75
  ) +
  geom_ribbon(
    data = prediction_lgi_age_trauma,
    mapping = aes(y = fit, ymin = lower, ymax = upper),
    alpha = 0.5,
    color = NA
  ) +
  scale_color_viridis_c(
    limits = c(0, 7),
    breaks = c(0, 3, 6)
  ) +
  scale_fill_viridis_c(
    limits = c(0, 7),
    breaks = c(0, 3, 6)
  ) +
  theme_linedraw() +
  theme(legend.position = "bottom") +
  labs(
    title = "LGI Developmental Trajectories by Trauma",
    subtitle = "lgi ~ s(age) + trauma + age:trauma",
    x = "Age",
    y = "LGI (global)",
    color = "Trauma",
    fill = "Trauma"
  )

plot_lgi_age_trauma
```

## Global LGI ~ SES * Sex {#two-ses-sex}
SES differed significantly by sex, with males having higher SES than females. Global LGI also differed significantly by sex after accounting for SES. The relationship between SES and global LGI was nonlinear, with evidence that the shape of this relationship differed by sex, although there was no evidence of a sex difference in the linear component of the SES association. The most complex model supported by the data therefore allowed the nonlinear association between SES and global LGI to vary by sex: lgi ~ sex + s(ses, by = sex). In this final model, SES was significantly associated with global LGI in both females (edf = 2.32, p < .001) and males (edf = 3.08, p < .001).
```{r}
#| label: lgi ses sex
data_lgi_ses_sex <-
  global_lgi |>
  inner_join(demo, by = "bblid") |>
  inner_join(env, by = "bblid") |>
  rename(ses = env_ses) |>
  select(lgi, sex, ses) |>
  mutate(
    sex_ord = ordered(sex, levels = c("Female", "Male")),
    ses_c = ses - mean(ses)
  ) |>
  drop_na()

t.test(
  ses ~ sex,
  data = data_lgi_ses_sex,
  var.equal = TRUE
)

plot_ses_sex <-
  ggplot(
    data = data_lgi_ses_sex,
    mapping = aes(x = ses, group = sex, color = sex, fill = sex)
  ) +
  geom_histogram(position = "identity", binwidth = 0.1, alpha = 0.5) +
  theme_linedraw() +
  theme(legend.position = "bottom") +
  color_scale_sex +
  fill_scale_sex +
  labs(title = "SES Distribution by Sex", x = "SES", y = "Count")

plot_ses_sex
```

### Global LGI ~ SES + Sex
```{r}
#| label: lgi ses sex 1
model_lgi_ses_sex_1 <- gam(
  lgi ~ ses + sex,
  data = data_lgi_ses_sex,
  method = "REML"
)

summary(model_lgi_ses_sex_1)
```

### Global LGI ~ SES * Sex
```{r}
#| label: lgi ses sex 2
model_lgi_ses_sex_2 <- gam(
  lgi ~ ses * sex,
  data = data_lgi_ses_sex,
  method = "REML"
)

summary(model_lgi_ses_sex_2)
```

```{r}
#| label: lgi ses sex 3
model_lgi_ses_sex_3 <- gam(
  lgi ~ ses_c + sex + s(ses_c, m = c(2, 0)),
  data = data_lgi_ses_sex,
  method = "REML"
)

summary(model_lgi_ses_sex_3)
```

```{r}
#| label: lgi ses sex 4
model_lgi_ses_sex_4 <- gam(
  lgi ~ ses_c * sex +
    s(ses_c, m = c(2, 0)) +
    s(ses_c, m = c(2, 0), by = sex_ord),
  data = data_lgi_ses_sex,
  method = "REML"
)

summary(model_lgi_ses_sex_4)
```

### Global LGI ~ SES * Sex (Final)
```{r}
#| label: lgi ses sex final
model_lgi_ses_sex_final <- gam(
  lgi ~ sex + s(ses, by = sex),
  data = data_lgi_ses_sex,
  method = "REML"
)

summary(model_lgi_ses_sex_final)

prediction_lgi_ses_sex <- data_lgi_ses_sex |>
  group_by(sex) |>
  summarise(
    ses = list(
      seq(
        min(ses, na.rm = TRUE),
        max(ses, na.rm = TRUE),
        length.out = 250
      )
    ),
    .groups = "drop"
  ) |>
  unnest(ses)

fit_lgi_ses_sex <-
  predict(
    model_lgi_ses_sex_final,
    newdata = prediction_lgi_ses_sex,
    type = "response",
    se.fit = TRUE
  ) |>
  map(as.vector) |>
  as_tibble() |>
  clean_names() |>
  mutate(lower = fit - 1.96 * se_fit, upper = fit + 1.96 * se_fit) |>
  select(fit, lower, upper)

prediction_lgi_ses_sex <- prediction_lgi_ses_sex |>
  bind_cols(fit_lgi_ses_sex)

plot_lgi_ses_sex <-
  ggplot(mapping = aes(x = ses, y = lgi, color = sex, fill = sex)) +
  geom_line(
    data = prediction_lgi_ses_sex,
    mapping = aes(y = fit),
    linewidth = 0.75
  ) +
  geom_ribbon(
    data = prediction_lgi_ses_sex,
    mapping = aes(y = fit, ymin = lower, ymax = upper),
    alpha = 0.5,
    color = NA
  ) +
  geom_point(
    data = data_lgi_ses_sex,
    position = position_jitterdodge(
      jitter.width = 0.5,
      dodge.width = 0.5
    ),
    alpha = 0.25,
    size = 1
  ) +
  theme_linedraw() +
  theme(legend.position = "bottom") +
  color_scale_sex +
  fill_scale_sex +
  labs(
    title = "SES LGI Relationship by Sex",
    subtitle = "lgi ~ ses + sex",
    x = "SES",
    y = "LGI (global)"
  )

plot_lgi_ses_sex
```

## Global LGI ~ Sex * Trauma {#two-sex-trauma}
Trauma exposure did not differ significantly between females and males. When trauma exposure and sex were modeled together, greater trauma exposure was significantly associated with lower global LGI, and males had significantly higher global LGI than females. There was no significant evidence that the trauma–LGI relationship was nonlinear, nor was there evidence that either the linear or nonlinear association between trauma exposure and global LGI differed by sex. The most complex model supported by the data was therefore `lgi ~ trauma + sex`.
```{r}
#| label: lgi ~ sex * trauma
data_lgi_sex_trauma <-
  global_lgi |>
  inner_join(demo, by = "bblid") |>
  inner_join(env, by = "bblid") |>
  rename(trauma = trauma_exposure) |>
  mutate(
    sex_ord = ordered(sex, levels = c("Female", "Male")),
    trauma_c = trauma - mean(trauma)
  ) |>
  select(lgi, sex, sex_ord, trauma, trauma_c) |>
  drop_na()

t.test(
  trauma ~ sex,
  data = data_lgi_sex_trauma,
  var.equal = TRUE
)
```

### Global LGI ~ Sex + Trauma + s(Trauma, m = c(2,0))
```{r}
#| label: lgi ~ sex + trauma + s(trauma, m = 0)
model_lgi_sex_trauma_1 <- gam(
  lgi ~ sex + trauma_c + s(trauma_c, k = 7, m = c(2, 0)),
  data = data_lgi_sex_trauma,
  method = "REML"
)

summary(model_lgi_sex_trauma_1)
```

### Global LGI ~ Sex * Trauma + s(Trauma, m = c(2,0))
```{r}
#| label: lgi ~ sex * trauma + s(trauma, m = 0)
model_lgi_sex_trauma_2 <- gam(
  lgi ~ sex * trauma_c + s(trauma_c, k = 7, m = c(2, 0)),
  data = data_lgi_sex_trauma,
  method = "REML"
)

summary(model_lgi_sex_trauma_2)
```

### Global LGI ~ Sex * Trauma + s(Trauma, m = c(2,0)) + s(Trauma, m = c(2,0), by = Sex)
```{r}
#| label: lgi ~ sex * trauma + s(trauma, m = 0) + s(trauma, m = 0, by = sex)
model_lgi_sex_trauma_3 <- gam(
  lgi ~ sex * trauma_c +
    s(trauma_c, k = 7, m = c(2, 0)) +
    s(trauma_c, k = 7, m = c(2, 0), by = sex_ord),
  data = data_lgi_sex_trauma,
  method = "REML"
)

summary(model_lgi_sex_trauma_3)
```

### Global LGI ~ Sex * Trauma
```{r}
#| label: lgi ~ sex * trauma [literal]
model_lgi_sex_trauma_4 <- gam(
  lgi ~ sex * trauma_c,
  data = data_lgi_sex_trauma,
  method = "REML"
)

summary(model_lgi_sex_trauma_4)
```

### Global LGI ~ Sex + Trauma
```{r}
#| label: lgi ~ sex + trauma
model_lgi_sex_trauma_5 <- gam(
  lgi ~ sex + trauma,
  data = data_lgi_sex_trauma,
  method = "REML"
)

summary(model_lgi_sex_trauma_5)
```


### Global LGI ~ Sex + Trauma (Final)
```{r}
#| label: lgi ~ sex + trauma [final]
model_lgi_sex_trauma_final <- model_lgi_sex_trauma_5

prediction_lgi_sex_trauma <- data_lgi_sex_trauma |>
  group_by(sex) |>
  summarise(
    trauma = list(
      seq(
        min(trauma, na.rm = TRUE),
        max(trauma, na.rm = TRUE),
        length.out = 250
      )
    ),
    .groups = "drop"
  ) |>
  unnest(trauma)

fit_data_lgi_sex_trauma <-
  predict(
    model_lgi_sex_trauma_final,
    newdata = prediction_lgi_sex_trauma,
    type = "response",
    se.fit = TRUE
  ) |>
  map(as.vector) |>
  as_tibble() |>
  clean_names() |>
  mutate(lower = fit - 1.96 * se_fit, upper = fit + 1.96 * se_fit) |>
  select(fit, lower, upper)

prediction_lgi_sex_trauma <- prediction_lgi_sex_trauma |>
  bind_cols(fit_data_lgi_sex_trauma)

plot_lgi_sex_trauma <-
  ggplot(mapping = aes(x = trauma, y = lgi, color = sex, fill = sex)) +
  geom_line(
    data = prediction_lgi_sex_trauma,
    mapping = aes(y = fit),
    linewidth = 0.75
  ) +
  geom_ribbon(
    data = prediction_lgi_sex_trauma,
    mapping = aes(y = fit, ymin = lower, ymax = upper),
    alpha = 0.5,
    color = NA
  ) +
  geom_point(
    data = data_lgi_sex_trauma,
    position = position_jitterdodge(
      jitter.width = 0.5,
      dodge.width = 0.5
    ),
    alpha = 0.25,
    size = 1
  ) +
  theme_linedraw() +
  theme(legend.position = "bottom") +
  color_scale_sex +
  fill_scale_sex +
  labs(
    title = "Trauma LGI Relationship by Sex",
    subtitle = "lgi ~ trauma + sex",
    x = "Trauma",
    y = "LGI (global)"
  )

plot_lgi_sex_trauma
```
