---
title: "Local Gyrification Index"
subtitle: "Exploratory Analyses"
author: "Zachary Heffernan"
date: today

format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true

execute: 
  echo: true
  warning: false
  message: false
  eval: true
---

```{r}
#| label: set up
#| include: false
# Import Libraries
library(tidyverse)
library(janitor)
library(mgcv)
library(kableExtra)
library(ComplexUpset)

# Source
source("R/plotting.R")
source("R/reporting.R")

# Define Paths
dir_dataset <- file.path(
  "/", "data", "secure", "bit_repo", "projects_repo", "ExtraLong", "data",
  "dataset_pnc_cross"
)

dir_evolpsy <- file.path(
  "/", "data", "secure", "bit_repo", "pi_repo", "raquel_gur", "evolpsy", "data"
)

path_imaging <- file.path(dir_dataset, "imaging_data.csv")
path_sub_ses <- file.path(dir_dataset, "subjects_sessions.csv")

path_merge <- file.path(
  dir_evolpsy, "go1_files", "pncgo1_data_merge_mar2025.csv"
)

path_env_ses <- file.path(dir_evolpsy, "zac_data", "env_scores.csv")
path_env_trauma <- file.path(dir_evolpsy, "zac_data", "env_for_zac.csv")

path_goassess <- file.path(
  "/", "data", "secure", "bit_repo", "projects_repo", "goassess_scoring",
  "results", "data_score.csv"
)

# Read Data
raw_imaging <- read_csv(path_imaging)
raw_sub_ses <- read_csv(path_sub_ses)

raw_merge <- read_csv(path_merge)

raw_env_ses <- read_csv(path_env_ses)
raw_env_trauma <- read_csv(path_env_trauma)

raw_goassess <- read_csv(path_goassess)
```

```{r}
#| label: clean data
#| include: false
# Demographics
demo <- raw_sub_ses |>
  select(bblid, age) |>
  left_join(
    select(raw_merge, bblid, sex, clinical_goassessDxpmr7),
    by = "bblid"
  ) |>
  rename(dx = clinical_goassessDxpmr7) |>
  mutate(
    age = age / 12,
    dx = str_to_lower(dx),
    sex = factor(
      case_when(sex == 1 ~ "Male", sex == 2 ~ "Female"),
      levels = c("Female", "Male")
    )
  )

# Environment
env <- full_join(raw_env_ses, raw_env_trauma, by = "bblid") |>
  select(bblid, envSES, Trauma_Exposure) |>
  clean_names()

# Imaging
imaging_qc <- raw_sub_ses |>
  mutate(qc = cume_dist(euler)) |>
  filter(qc > 0.01) |>
  select(bblid, scanid)

imaging <- raw_imaging |>
  filter(atlas == "desikan_killiany", !is.na(lobe)) |>
  semi_join(imaging_qc, by = c("bblid", "scanid")) |>
  pivot_wider(values_from = value, names_from = metric) |>
  select(bblid, scanid, laterality, region, lgi, area)

global_lgi <- imaging |>
  group_by(bblid, scanid) |>
  summarize(lgi = weighted.mean(lgi, area), .groups = "drop")

# Clinical
clinical <- raw_merge |>
  select(bblid, starts_with("clinical_longfactor_corrTraits_")) |>
  pivot_longer(cols = -bblid, values_to = "score", names_to = "factor") |>
  mutate(factor = str_replace(factor, "clinical_longfactor_corrTraits_", ""))

# Cognitive
cnb <- raw_merge |>
  select(bblid, starts_with("cnb") & ends_with("efficiency_agereg")) |>
  pivot_longer(cols = -bblid, values_to = "efficiency", names_to = "domain") |>
  mutate(
    domain = str_replace_all(domain, c("cnb_" = "", "_efficiency_agereg" = ""))
  ) |>
  filter(domain != "discounting")

# GOASSESS
goassess <-
  raw_goassess |>
  semi_join(demo, by = c("subject_id" = "bblid")) |>
  filter(version == 1) |>
  group_by(subject_id) |>
  summarize(across(starts_with("smry_"), \(x) max(x, na.rm = TRUE)))
```

# Sample Description
## Demographics Table
```{r}
#| label: demos table

demo_table <- demo |>
  summarize(
    n = n(),
    age_mean = mean(age, na.rm = TRUE),
    age_sd = sd(age, na.rm = TRUE),
    male_n = sum(sex == "Male"),
    female_n = sum(sex == "Female"),
    dx_td_n = sum(dx == "td", na.rm = TRUE),
    dx_ps_n = sum(dx == "ps", na.rm = TRUE),
    dx_op_n = sum(dx == "op", na.rm = TRUE)
  ) |>
  mutate(
    age = sprintf("%.2f (%.2f)", age_mean, age_sd),
    male = sprintf("%d (%.2f%%)", male_n, (male_n * 100 / n)),
    female = sprintf("%d (%.2f%%)", female_n, (female_n * 100 / n)),
    td = sprintf("%d (%.2f%%)", dx_td_n, (dx_td_n * 100 / n)),
    ps = sprintf("%d (%.2f%%)", dx_ps_n, (dx_ps_n * 100 / n)),
    op = sprintf("%d (%.2f%%)", dx_op_n, (dx_op_n * 100 / n))
  ) |>
  mutate(n = as.character(n)) |>
  select(n, age, male, female, td, ps, op) |>
  pivot_longer(
    cols = everything(),
    names_to = "Characteristic",
    values_to = "Value"
  ) |>
  mutate(Category = c("n", "age", "sex", "sex", "dx", "dx", "dx"))

row_hierarchy <- demo_table |>
  mutate(row_number = row_number()) |>
  group_by(Category) |>
  summarise(
    start = min(row_number),
    end = max(row_number),
    .groups = "drop"
  ) |>
  filter(start != end)

demo_kable <- demo_table |>
  select(-Category) |>
  kbl(
    caption = "PNC Cross Sectional Demographics",
    format = "html",
    col.names = c("", "Overall"),
    align = c("l", "c")
  ) |>
  kable_styling(
    full_width = FALSE,
    position = "left",
    font_size = 12
  )

for (i in seq_len(nrow(row_hierarchy))) {
  demo_kable <- demo_kable |>
    group_rows(
      group = row_hierarchy$Category[i],
      start_row = row_hierarchy$start[i],
      end_row = row_hierarchy$end[i]
    )
}

demo_kable
```

## Comorbidities
```{r}
#| label: comorbidities
comorbid <- goassess |>
  select(subject_id, ends_with("cat")) |>
  rename(bblid = subject_id) |>
  rename_with(\(x) str_replace_all(x, c("smry_" = "", "_cat" = ""))) |>
  mutate(across(
    -bblid,
    \(x) case_when(x %in% c(0, 1, 2, 3) ~ FALSE, x == 4 ~ TRUE)
  )) |>
  left_join(global_lgi, by = "bblid") |>
  select(-bblid, -scanid) |>
  drop_na()

sets <- setdiff(names(comorbid), "lgi")

psy_intersections <- unlist(
  lapply(
    seq_along(sets),
    \(n) combn(sets, n, simplify = FALSE)
  ),
  recursive = FALSE
) |>
  keep(\(x) "psy" %in% x)

psy_queries <- map(
  psy_intersections,
  \(x) {
    upset_query(
      intersect = x,
      color = "red",
      fill = "red"
    )
  }
)

plot_comorbid <- upset(
  comorbid,
  sets,
  base_annotations = list(
    "Intersection size" = intersection_size(
      counts = FALSE
    )
  ),
  annotations = list(
    "LGI" = ggplot(mapping = aes(x = intersection, y = lgi, color = psy)) +
      geom_boxplot() +
      scale_color_manual(
        values = c(`TRUE` = "red", `FALSE` = "black"),
        guide = "none"
      )
  ),
  queries = psy_queries
)

plot_comorbid
```

# Local Gyrification Index
## LGI vs eTIV
```{r}
#| label: lgi vs etiv
data_etiv <- raw_imaging |>
  filter(region == "EstimatedTotalIntraCranialVol") |>
  rename(etiv = value) |>
  select(bblid, scanid, etiv)

data_lgi_etiv <- left_join(global_lgi, data_etiv, by = c("bblid", "scanid"))

plot_lgi_etiv <-
  ggplot(data = data_lgi_etiv, mapping = aes(x = etiv, y = lgi)) +
  geom_point() +
  theme_linedraw()

plot_lgi_etiv
```

# Main Effects
  - [There is a significant nonlinear effect of Age on Global LGI, where Global LGI declines with Age.](#main-age)
  - [There is a significant positive correlation between Socioeconomic Status and Global LGI.](#main-ses)
  - [There is a significant nonlinear effect of Trauma Exposure on Global LGI, where those with higher Trauma Exposure have lower Global LGI.](#main-trauma)
  - [There is a significant difference in Global LGI between sexes, with Males having higher Global LGI than Females.](#main-sex)
  - [There is a significant decreasing trend in Global LGI across increasing levels of GOASSESS Psychosis Scores.](#main-goa-psy)

## Global LGI ~ Age {#main-age}
Global LGI is significantly negatively correlated with age. This effect is significantly nonlinear, though the nonlinearity is subtle.
```{r}
#| label: lgi ~ age
data_lgi_age <-
  inner_join(global_lgi, demo, by = c("bblid")) |>
  select(lgi, age) |>
  drop_na()

test_lgi_age <-
  cor.test(data_lgi_age$lgi, data_lgi_age$age)

plot_lgi_age <-
  ggplot(data = data_lgi_age, aes(x = age, y = lgi)) +
  geom_point() +
  theme_linedraw() +
  labs(
    title = "Correlation of Global LGI and Age",
    subtitle = report_correlation(test_lgi_age),
    x = "Age",
    y = "LGI (global)"
  )

plot_lgi_age
```

### Global LGI ~ s(Age)
```{r}
#| label: lgi ~ s(age)
model_lgi_age <- gam(
  lgi ~ age + s(age, m = c(2, 0)),
  data = data_lgi_age,
  method = "REML"
)

summary(model_lgi_age)

plot_lgi_age_linearity <- ggplot(data = data_lgi_age, aes(x = age, y = lgi)) +
  geom_point() +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = FALSE,
    aes(color = "Linear")
  ) +
  geom_smooth(
    method = "gam",
    formula = y ~ s(x),
    se = FALSE,
    aes(color = "Nonlinear")
  ) +
  scale_color_manual(
    values = c("Linear" = "blue", "Nonlinear" = "red"),
    name = NULL
  ) +
  theme_linedraw() +
  theme(legend.position = "bottom") +
  labs(
    title = "Linearity of Relationship Between Global LGI and Age",
    x = "Age",
    y = "LGI (global)"
  )

plot_lgi_age_linearity
```

## Global LGI ~ Socioeconomic Status {#main-ses}
Global LGI is signitifantly positively correlated with socioeconomic status. This effect is not significantly nonlinear.
```{r}
#| label: lgi ~ ses
data_lgi_ses <-
  inner_join(global_lgi, env, by = "bblid") |>
  select(lgi, env_ses) |>
  drop_na()

test_lgi_ses <-
  cor.test(data_lgi_ses$lgi, data_lgi_ses$env_ses)

plot_lgi_ses <-
  ggplot(data = data_lgi_ses, aes(x = env_ses, y = lgi)) +
  geom_point() +
  theme_linedraw() +
  labs(
    title = "Correlation of Global LGI and Socioeconomic Status",
    subtitle = report_correlation(test_lgi_ses),
    x = "Socioeconomic Status",
    y = "LGI (global)"
  )

plot_lgi_ses
```

### Global LGI ~ s(Socioeconomic Status)
```{r}
#| label: lgi ~ s(ses)
model_lgi_ses <- gam(
  lgi ~ env_ses + s(env_ses, m = c(2, 0)),
  data = data_lgi_ses,
  method = "REML"
)

summary(model_lgi_ses)
```

## Global LGI ~ Trauma Exposure {#main-trauma}
Global LGI is significantly negatively correlated with Trauma Exposure. This effect is significantly nonlinear.
```{r}
#| label: lgi ~ trauma
data_lgi_trauma <-
  inner_join(global_lgi, env, by = "bblid") |>
  select(lgi, trauma_exposure) |>
  drop_na()

test_lgi_trauma <-
  cor.test(data_lgi_trauma$lgi, data_lgi_trauma$trauma_exposure)

plot_lgi_trauma <-
  ggplot(data = data_lgi_trauma, aes(x = trauma_exposure, y = lgi)) +
  geom_point() +
  theme_linedraw() +
  labs(
    title = "Correlation of Global LGI and Trauma Exposure",
    subtitle = report_correlation(test_lgi_trauma),
    x = "Trauma Exposure",
    y = "LGI (global)"
  )

plot_lgi_trauma
```

### Global LGI ~ s(Trauma Exposure)
```{r}
#| label: lgi ~ s(trauma)
model_lgi_trauma <- gam(
  lgi ~ trauma_exposure + s(trauma_exposure, k = 7, m = c(2, 0)),
  data = data_lgi_trauma,
  method = "REML"
)

summary(model_lgi_trauma)

plot_lgi_trauma_linearity <-
  ggplot(data = data_lgi_trauma, aes(x = trauma_exposure, y = lgi)) +
  geom_point() +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = FALSE,
    aes(color = "Linear")
  ) +
  geom_smooth(
    method = "gam",
    formula = y ~ s(x, k = 7),
    se = FALSE,
    aes(color = "Nonlinear")
  ) +
  scale_color_manual(
    values = c("Linear" = "blue", "Nonlinear" = "red"),
    name = NULL
  ) +
  theme_linedraw() +
  theme(legend.position = "bottom") +
  labs(
    title = "Linearity of Relationship Between Global LGI and Trauma Exposure",
    x = "Trauma Exposure",
    y = "LGI (global)"
  )

plot_lgi_trauma_linearity
```

## Global LGI ~ Sex {#main-sex}
There are significant differences in Global LGI between sexes, with Males having higher Global LGI than Females.
```{r}
#| label: lgi ~ sex
data_lgi_sex <-
  inner_join(global_lgi, demo, by = "bblid") |>
  select(lgi, sex) |>
  drop_na()

test_lgi_sex <-
  t.test(lgi ~ sex, data = data_lgi_sex)

plot_lgi_sex <-
  ggplot(data = data_lgi_sex, aes(x = sex, y = lgi, color = sex, fill = sex)) +
  geom_jitter(alpha = 0.25) +
  geom_violin(alpha = 0.5) +
  theme_linedraw() +
  theme(legend.position = "none") +
  color_scale_sex +
  fill_scale_sex +
  labs(
    title = "Differences in Global LGI between Sexes",
    subtitle = report_t_test(test_lgi_sex),
    x = "Sex",
    y = "LGI (global)"
  )

plot_lgi_sex
```

## Global LGI ~ GOASSESS Psychosis Category {#main-goa-psy}
Global LGI shows a significant decreasing trend across increasing levels of GOASSESS Psychosis Scores.
```{r}
#| label: lgi ~ goa_smry_psy
data_lgi_psy <-
  inner_join(global_lgi, goassess, by = c("bblid" = "subject_id")) |>
  select(lgi, smry_psy_cat) |>
  filter(smry_psy_cat %in% c(0, 1, 2, 3, 4)) |>
  mutate(smry_psy_cat = ordered(smry_psy_cat)) |>
  drop_na()

test_lgi_psy <-
  clinfun::jonckheere.test(
    x = data_lgi_psy$lgi,
    g = data_lgi_psy$smry_psy_cat,
    alternative = "decreasing"
  )

plot_lgi_psy <-
  ggplot(
    data = data_lgi_psy,
    aes(x = smry_psy_cat, y = lgi, color = smry_psy_cat, fill = smry_psy_cat)
  ) +
  geom_jitter(alpha = 0.25) +
  geom_violin(alpha = 0.5) +
  theme_linedraw() +
  theme(legend.position = "none") +
  scale_x_discrete(
    labels = c(
      "0" = "Not Endorsed",
      "1" = "Endorsed",
      "2" = "Subthreshold",
      "3" = "Threshold",
      "4" = "Significant"
    )
  ) +
  labs(
    title = "Differences in Global LGI between Psychosis Categories",
    subtitle = report_jt(test_lgi_psy),
    x = "GOASSESS Psychosis Category",
    y = "LGI (global)"
  )

plot_lgi_psy
```

# Two-Way Interactions
  - [There is a significant difference in age between sexes, with females being older than males. There are no significant linear or nonlinear interactions between age and sex. The most complex model supported is `lgi ~ s(age) + sex`.](#two-age-sex)
  - [Global LGI differed by sex and showed a significant nonlinear association with SES, with evidence that the shape of this SES–LGI relationship differed between females and males. The final supported model is `lgi ~ s(ses, by = sex) + sex`.](#two-ses-sex)
  - [There is no significant difference in trauma between sexes. There are no significant linear or nonlinear interactions between sex and trauma. When sex and trauma are both included in the model, trauma is no longer significantly nonlinear like it is when in the model by itself. The most complex model supported is `lgi ~ sex + trauma`.](#two-sex-trauma)

## Global LGI ~ Age * Sex {#two-age-sex}
There are no significant linear or nonlinear interactions between age and sex. The most complex model supported is `lgi ~ s(age) + sex`
```{r}
#| label: lgi ~ age * sex
data_lgi_age_sex <-
  inner_join(global_lgi, demo, by = c("bblid")) |>
  mutate(
    sex_ord = ordered(sex, levels = c("Female", "Male")),
    age_c = age - mean(age)
  ) |>
  select(lgi, age, age_c, sex, sex_ord) |>
  drop_na()

t.test(
  age ~ sex,
  data = data_lgi_age_sex,
  var.equal = TRUE
)

plot_age_sex <-
  ggplot(
    data = data_lgi_age_sex,
    mapping = aes(x = age, group = sex, color = sex, fill = sex)
  ) +
  geom_histogram(position = "identity", binwidth = 1, alpha = 0.5) +
  theme_linedraw() +
  theme(legend.position = "bottom") +
  color_scale_sex +
  fill_scale_sex +
  labs(title = "Age Distribution by Sex", x = "Age", y = "Count")

plot_age_sex
```

### Global LGI ~ Age + Sex + s(Age, m = c(2,0))
Sex contributes when accounting for age. Age contributes when accounting for sex, and this effect is significantly nonlinear.
```{r}
#| label: lgi ~ age + sex + s(age, m = 0)
model_lgi_age_sex_1 <- gam(
  lgi ~ age_c + sex + s(age_c, m = c(2, 0)),
  data = data_lgi_age_sex,
  method = "REML"
)

summary(model_lgi_age_sex_1)
```

### Global LGI ~ Sex * Age + s(Age, m = c(2,0))
Age and sex do not have a significant linear interaction.
```{r}
#| label: lgi ~ age * sex + s(age, m = 0)
model_lgi_age_sex_2 <- gam(
  lgi ~ age_c * sex + s(age_c, m = c(2, 0)),
  data = data_lgi_age_sex,
  method = "REML"
)

summary(model_lgi_age_sex_2)
```

### Global LGI ~ Sex * Age + s(Age, m = c(2,0)) + s(Age, m = c(2,0), by = Sex)
Age and sex do not have a significant nonlinear interaction.
```{r}
#| label: lgi ~ age * sex + s(age, m = 0) + s(age, m = 0, by = sex)
model_lgi_age_sex_3 <- gam(
  lgi ~ age_c * sex +
    s(age_c, m = c(2, 0)) +
    s(age_c, m = c(2, 0), by = sex_ord),
  data = data_lgi_age_sex,
  method = "REML"
)

summary(model_lgi_age_sex_3)
```

### Global LGI ~ Age * Sex (Final)
```{r}
#| label: lgi ~ age * sex [final]
model_lgi_age_sex_final <- gam(
  lgi ~ sex + s(age),
  data = data_lgi_age_sex,
  method = "REML"
)

prediction_lgi_age_sex <- data_lgi_age_sex |>
  group_by(sex) |>
  summarise(
    age = list(
      seq(
        min(age, na.rm = TRUE),
        max(age, na.rm = TRUE),
        length.out = 250
      )
    ),
    .groups = "drop"
  ) |>
  unnest(age)

fit_data_lgi_age_sex <- predict(
  model_lgi_age_sex_final,
  newdata = prediction_lgi_age_sex,
  type = "response",
  se.fit = TRUE
)

prediction_lgi_age_sex <- prediction_lgi_age_sex |>
  mutate(
    fit = as.vector(pluck(fit_data_lgi_age_sex, "fit")),
    se_fit = as.vector(pluck(fit_data_lgi_age_sex, "se.fit")),
    lower = fit - 1.96 * se_fit,
    upper = fit + 1.96 * se_fit
  ) |>
  select(age, sex, fit, upper, lower)

plot_lgi_age_sex <-
  ggplot(mapping = aes(x = age, y = lgi, color = sex, fill = sex)) +
  geom_point(data = data_lgi_age_sex, alpha = 0.25, size = 1) +
  geom_line(
    data = prediction_lgi_age_sex,
    mapping = aes(y = fit),
    linewidth = 0.75
  ) +
  geom_ribbon(
    data = prediction_lgi_age_sex,
    mapping = aes(y = fit, ymin = lower, ymax = upper),
    alpha = 0.5,
    color = NA
  ) +
  theme_linedraw() +
  theme(legend.position = "bottom") +
  color_scale_sex +
  fill_scale_sex +
  labs(
    title = "LGI Developmental Trajectories by Sex",
    subtitle = "lgi ~ s(age) + sex",
    x = "Age",
    y = "LGI (global)"
  )

plot_lgi_age_sex
```

## Global LGI ~ SES * Sex {#two-ses-sex}
SES differed significantly by sex, with males having higher SES than females. Global LGI also differed significantly by sex after accounting for SES. The relationship between SES and global LGI was nonlinear, with evidence that the shape of this relationship differed by sex, although there was no evidence of a sex difference in the linear component of the SES association. The most complex model supported by the data therefore allowed the nonlinear association between SES and global LGI to vary by sex: lgi ~ sex + s(ses, by = sex). In this final model, SES was significantly associated with global LGI in both females (edf = 2.32, p < .001) and males (edf = 3.08, p < .001).
```{r}
#| label: lgi ~ sex * sex
data_lgi_ses_sex <-
  global_lgi |>
  inner_join(demo, by = "bblid") |>
  inner_join(env, by = "bblid") |>
  rename(ses = env_ses) |>
  select(lgi, sex, ses) |>
  mutate(
    sex_ord = ordered(sex, levels = c("Female", "Male")),
    ses_c = ses - mean(ses)
  ) |>
  drop_na()

t.test(
  ses ~ sex,
  data = data_lgi_ses_sex,
  var.equal = TRUE
)

plot_ses_sex <-
  ggplot(
    data = data_lgi_ses_sex,
    mapping = aes(x = ses, group = sex, color = sex, fill = sex)
  ) +
  geom_histogram(position = "identity", binwidth = 0.1, alpha = 0.5) +
  theme_linedraw() +
  theme(legend.position = "bottom") +
  color_scale_sex +
  fill_scale_sex +
  labs(title = "SES Distribution by Sex", x = "SES", y = "Count")

plot_ses_sex
```

### Global LGI ~ SES + Sex
```{r}
#| label: lgi ~ ses + sex
model_lgi_ses_sex_1 <- gam(
  lgi ~ ses + sex,
  data = data_lgi_ses_sex,
  method = "REML"
)

summary(model_lgi_ses_sex_1)
```

### Global LGI ~ SES * Sex
```{r}
#| label: lgi ~ ses * sex [literal]
model_lgi_ses_sex_2 <- gam(
  lgi ~ ses * sex,
  data = data_lgi_ses_sex,
  method = "REML"
)

summary(model_lgi_ses_sex_2)
```

```{r}
model_lgi_ses_sex_3 <- gam(
  lgi ~ ses_c + sex + s(ses_c, m = c(2, 0)),
  data = data_lgi_ses_sex,
  method = "REML"
)

summary(model_lgi_ses_sex_3)
```

```{r}
model_lgi_ses_sex_4 <- gam(
  lgi ~ ses_c * sex +
    s(ses_c, m = c(2, 0)) +
    s(ses_c, m = c(2, 0), by = sex_ord),
  data = data_lgi_ses_sex,
  method = "REML"
)

summary(model_lgi_ses_sex_4)
```

### Global LGI ~ SES * Sex (Final)
```{r}
#| label: lgi ~ ses * sex [final]
model_lgi_ses_sex_final <- gam(
  lgi ~ sex + s(ses, by = sex),
  data = data_lgi_ses_sex,
  method = "REML"
)

summary(model_lgi_ses_sex_final)

prediction_lgi_ses_sex <- data_lgi_ses_sex |>
  group_by(sex) |>
  summarise(
    ses = list(
      seq(
        min(ses, na.rm = TRUE),
        max(ses, na.rm = TRUE),
        length.out = 250
      )
    ),
    .groups = "drop"
  ) |>
  unnest(ses)

fit_lgi_ses_sex <-
  predict(
    model_lgi_ses_sex_final,
    newdata = prediction_lgi_ses_sex,
    type = "response",
    se.fit = TRUE
  ) |>
  map(as.vector) |>
  as_tibble() |>
  clean_names() |>
  mutate(lower = fit - 1.96 * se_fit, upper = fit + 1.96 * se_fit) |>
  select(fit, lower, upper)

prediction_lgi_ses_sex <- prediction_lgi_ses_sex |>
  bind_cols(fit_lgi_ses_sex)

plot_lgi_ses_sex <-
  ggplot(mapping = aes(x = ses, y = lgi, color = sex, fill = sex)) +
  geom_line(
    data = prediction_lgi_ses_sex,
    mapping = aes(y = fit),
    linewidth = 0.75
  ) +
  geom_ribbon(
    data = prediction_lgi_ses_sex,
    mapping = aes(y = fit, ymin = lower, ymax = upper),
    alpha = 0.5,
    color = NA
  ) +
  geom_point(
    data = data_lgi_ses_sex,
    position = position_jitterdodge(
      jitter.width = 0.5,
      dodge.width = 0.5
    ),
    alpha = 0.25,
    size = 1
  ) +
  theme_linedraw() +
  theme(legend.position = "bottom") +
  color_scale_sex +
  fill_scale_sex +
  labs(
    title = "SES LGI Relationship by Sex",
    subtitle = "lgi ~ ses + sex",
    x = "SES",
    y = "LGI (global)"
  )

plot_lgi_ses_sex
```

## Global LGI ~ Sex * Trauma {#two-sex-trauma}
```{r}
#| label: lgi ~ sex * trauma
data_lgi_sex_trauma <-
  global_lgi |>
  inner_join(demo, by = "bblid") |>
  inner_join(env, by = "bblid") |>
  rename(trauma = trauma_exposure) |>
  mutate(
    sex_ord = ordered(sex, levels = c("Female", "Male")),
    trauma_c = trauma - mean(trauma)
  ) |>
  select(lgi, sex, sex_ord, trauma, trauma_c) |>
  drop_na()

t.test(
  trauma ~ sex,
  data = data_lgi_sex_trauma,
  var.equal = TRUE
)
```

### Global LGI ~ Sex + Trauma + s(Trauma, m = c(2,0))
```{r}
#| label: lgi ~ sex + trauma + s(trauma, m = 0)
model_lgi_sex_trauma_1 <- gam(
  lgi ~ sex + trauma_c + s(trauma_c, k = 7, m = c(2, 0)),
  data = data_lgi_sex_trauma,
  method = "REML"
)

summary(model_lgi_sex_trauma_1)
```

### Global LGI ~ Sex * Trauma + s(Trauma, m = c(2,0))
```{r}
#| label: lgi ~ sex * trauma + s(trauma, m = 0)
model_lgi_sex_trauma_2 <- gam(
  lgi ~ sex * trauma_c + s(trauma_c, k = 7, m = c(2, 0)),
  data = data_lgi_sex_trauma,
  method = "REML"
)

summary(model_lgi_sex_trauma_2)
```

### Global LGI ~ Sex * Trauma + s(Trauma, m = c(2,0)) + s(Trauma, m = c(2,0), by = Sex)
```{r}
#| label: lgi ~ sex * trauma + s(trauma, m = 0) + s(trauma, m = 0, by = sex)
model_lgi_sex_trauma_3 <- gam(
  lgi ~ sex * trauma_c +
    s(trauma_c, k = 7, m = c(2, 0)) +
    s(trauma_c, k = 7, m = c(2, 0), by = sex_ord),
  data = data_lgi_sex_trauma,
  method = "REML"
)

summary(model_lgi_sex_trauma_3)
```

### Global LGI ~ Sex * Trauma
```{r}
#| label: lgi ~ sex * trauma [literal]
model_lgi_sex_trauma_4 <- gam(
  lgi ~ sex * trauma_c,
  data = data_lgi_sex_trauma,
  method = "REML"
)

summary(model_lgi_sex_trauma_4)
```

### Global LGI ~ Sex + Trauma
```{r}
#| label: lgi ~ sex + trauma
model_lgi_sex_trauma_5 <- gam(
  lgi ~ sex + trauma,
  data = data_lgi_sex_trauma,
  method = "REML"
)

summary(model_lgi_sex_trauma_5)
```


### Global LGI ~ Sex + Trauma (Final)
```{r}
#| label: lgi ~ sex + trauma [final]
model_lgi_sex_trauma_final <- model_lgi_sex_trauma_5

prediction_lgi_sex_trauma <- data_lgi_sex_trauma |>
  group_by(sex) |>
  summarise(
    trauma = list(
      seq(
        min(trauma, na.rm = TRUE),
        max(trauma, na.rm = TRUE),
        length.out = 250
      )
    ),
    .groups = "drop"
  ) |>
  unnest(trauma)

fit_data_lgi_sex_trauma <-
  predict(
    model_lgi_sex_trauma_final,
    newdata = prediction_lgi_sex_trauma,
    type = "response",
    se.fit = TRUE
  ) |>
  map(as.vector) |>
  as_tibble() |>
  clean_names() |>
  mutate(lower = fit - 1.96 * se_fit, upper = fit + 1.96 * se_fit) |>
  select(fit, lower, upper)

prediction_lgi_sex_trauma <- prediction_lgi_sex_trauma |>
  bind_cols(fit_data_lgi_sex_trauma)

plot_lgi_sex_trauma <-
  ggplot(mapping = aes(x = trauma, y = lgi, color = sex, fill = sex)) +
  geom_line(
    data = prediction_lgi_sex_trauma,
    mapping = aes(y = fit),
    linewidth = 0.75
  ) +
  geom_ribbon(
    data = prediction_lgi_sex_trauma,
    mapping = aes(y = fit, ymin = lower, ymax = upper),
    alpha = 0.5,
    color = NA
  ) +
  geom_point(
    data = data_lgi_sex_trauma,
    position = position_jitterdodge(
      jitter.width = 0.5,
      dodge.width = 0.5
    ),
    alpha = 0.25,
    size = 1
  ) +
  theme_linedraw() +
  theme(legend.position = "bottom") +
  color_scale_sex +
  fill_scale_sex +
  labs(
    title = "Trauma LGI Relationship by Sex",
    subtitle = "lgi ~ trauma + sex",
    x = "Trauma",
    y = "LGI (global)"
  )

plot_lgi_sex_trauma
```
